---
title: "Group Exercise: Geostatistical Analysis of Aluminium (Al) in the KOLA Data Set"
subtitle: "VU Environmental Statistics SS2026 — Group 2 Homework"
author: "Lehmann Leon, /n Hovhannisyan Norayr, /n Strachinaru Andrei"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options: show code, suppress warnings/messages for cleaner output
knitr::opts_chunk$set(
  echo    = TRUE,      # Show R code in the report
  warning = FALSE,     # Suppress warnings (e.g. from geoR)
  message = FALSE,     # Suppress package loading messages
  fig.align = "center",
  fig.width  = 7,
  fig.height = 5
)
options(digits = 4)
```

# Introduction

This report presents a geostatistical analysis of the element **Aluminium (Al)** from the
KOLA-95 geochemical data set (`KOLA95_Chor.txt`). The data originate from a systematic
sampling campaign across the Kola Peninsula (Russia/Finland/Norway) and contain
concentrations of numerous chemical elements in O-horizon (organic) soil samples.

The analysis follows these steps:

1. Data import
2. Explorative statistics — choosing a data transformation
3. Spatial plot of Al values
4. Variogram cloud and empirical variogram
5. Fitting a theoretical variogram model
6. Spatial prediction (kriging) and mapping
7. Voluntary: sensitivity of the kriging map to the variogram model

---

# Data Import

```{r data-import}
# Load the geoR package, which provides all geostatistical tools used below.
# install.packages("geoR")  # run once if not yet installed
library(geoR)

# Read the KOLA data from a tab-separated text file with a header row.
# The file must be in the same directory as this Rmd file.
KOLA95 <- read.table(
  "KOLA95_Chor.txt",
  sep    = "\t",   # columns are separated by tabs
  header = TRUE    # first row contains column names
)

# Identify the column index of "Al" to use in as.geodata()
al_col <- which(names(KOLA95) == "Al")
cat("Column index of Al:", al_col, "\n")
cat("Total number of samples:", nrow(KOLA95), "\n")
```

```{r geodata}
# Convert the data frame to a geoR "geodata" object.
# coords.col = 2:3 uses columns XCOO and YCOO as spatial coordinates (UTM metres).
# data.col   = al_col selects the Al column as the target variable.
s100 <- as.geodata(
  KOLA95,
  coords.col = 2:3,          # XCOO (col 2) and YCOO (col 3)
  data.col   = al_col        # Al concentrations (single column — no data.names needed)
)

# Quick summary of the geodata object: spatial extent + data statistics
summary(s100)
```

*The summary shows the spatial extent of the sampling grid and a first overview of Al
concentrations. Note the large range between minimum and maximum, which is a first hint
that a log-transformation may be appropriate.*

---

# Explorative Statistics — Choosing a Transformation {#exploration}

Before computing variograms we check whether the data are approximately **symmetric
(Gaussian)**, as classical kriging assumes normality. We compare the histogram of the
original values with that of the log-transformed values.

```{r hist-original, fig.cap="Histogram of original Al concentrations. Strong right-skew suggests a log-transformation is advisable."}
# Histogram of the raw Al concentrations
hist(
  s100$data,
  breaks = 30,
  main   = "Histogram of Al (original scale)",
  xlab   = "Al concentration (mg/kg)",
  col    = "steelblue",
  border = "white"
)
# Add a vertical line at the mean to illustrate the skewness
abline(v = mean(s100$data), col = "red", lwd = 2, lty = 2)
legend("topright", legend = paste("Mean =", round(mean(s100$data), 1)),
       col = "red", lty = 2, lwd = 2, cex = 0.9)
```

```{r hist-log, fig.cap="Histogram of log-transformed Al concentrations. The distribution is much more symmetric, justifying the log-transformation."}
# Log-transform the Al values and inspect the resulting distribution
hist(
  log(s100$data),
  breaks = 30,
  main   = "Histogram of log(Al)",
  xlab   = "log(Al concentration)",
  col    = "darkorange",
  border = "white"
)
abline(v = mean(log(s100$data)), col = "red", lwd = 2, lty = 2)
legend("topright", legend = paste("Mean =", round(mean(log(s100$data)), 3)),
       col = "red", lty = 2, lwd = 2, cex = 0.9)
```

```{r apply-log}
# Decision: the log-transformed values are approximately symmetric (bell-shaped),
# while the original values are strongly right-skewed.
# --> We apply the log-transformation before geostatistical modelling.
s100$data <- log(s100$data)

# Verify the transformation was applied correctly
cat("Summary of log(Al):\n")
print(summary(s100$data))
```

**Interpretation:** The original Al concentrations are strongly right-skewed (long upper
tail), which would violate the normality assumption of ordinary kriging. After
log-transformation the histogram is approximately bell-shaped, making the data suitable
for geostatistical analysis. All further steps are performed on **log(Al)**.

---

# Spatial Plot of Al Values {#spatialplot}

We plot the sample locations coloured by their log(Al) value to get a first impression
of the spatial structure (e.g. spatial gradients, clusters of high/low values).

```{r spatial-plot, fig.cap="Sample locations coloured by log(Al) quartile. Blue: low; Green: medium-low; Yellow: medium-high; Red: high concentrations. A spatial gradient is visible from NW (lower values) to SE (higher values)."}
# plot.geodata() produces a 4-panel plot:
# top-left: data vs X-coordinate; top-right: data locations (coloured by quartile);
# bottom-left: data locations; bottom-right: data vs Y-coordinate.
plot(s100,
     lowess = TRUE   # adds a smooth lowess trend line to the marginal plots
)
```

```{r points-plot, fig.cap="Spatial distribution of log(Al) values. Symbol size and colour reflect the quartile class (blue=Q1, green=Q2, yellow=Q3, red=Q4)."}
# A cleaner map showing only the spatial locations coloured by quartile
points(s100,
       xlab = "Easting (m)",
       ylab = "Northing (m)",
       main = "Sample locations — log(Al) by quartile"
)
```

**Interpretation:** The plot reveals a broad-scale spatial structure: samples in the
south-eastern part of the study area tend to have higher Al concentrations (red/yellow
symbols), while north-western samples show lower values (blue/green). This spatial
pattern motivates a formal variogram analysis to quantify the spatial autocorrelation.

---

# Empirical Variogram Analysis {#variogram}

## Variogram Cloud

The **variogram cloud** plots the squared difference of all sample pairs
$\frac{1}{2}[z(x_i) - z(x_j)]^2$ against their separation distance $h$. It gives an
initial impression of spatial variability and helps detect outliers.

```{r variogram-cloud, fig.cap="Variogram cloud of log(Al). Each point represents one sample pair. The increasing trend with distance indicates positive spatial autocorrelation."}
# Compute and plot the omnidirectional variogram cloud.
# max.dist = 400000 m (400 km) — covers roughly 2/3 of the study area's extent,
# which is a common rule of thumb to avoid unreliable estimates at large lags.
cloud1 <- variog(s100, option = "cloud", max.dist = 400000)
plot(cloud1,
     main = "Variogram cloud — log(Al)",
     xlab = "Distance (m)",
     ylab = expression(frac(1, 2) * "[z(x[i]) - z(x[j])]"^2)
)
```

**Interpretation:** The variogram cloud shows an increasing trend in semivariance with
distance, which confirms positive spatial autocorrelation — nearby locations tend to
have more similar Al concentrations than distant ones. The scatter is large, as expected
for a cloud, but the upward trend is clear.

## Empirical (Binned) Variogram

We bin the pairs by distance to obtain a more readable **experimental variogram**.

```{r empirical-variogram, fig.cap="Experimental (binned) variogram of log(Al) with 10 lag classes up to 400 km. The semivariance increases and then levels off, indicating the range of spatial dependence."}
# Bin the variogram into 10 equal-width lag classes from 0 to 400,000 m.
# uvec defines the bin boundaries; l=11 creates 10 intervals.
bin1 <- variog(s100, uvec = seq(0, 400000, l = 11))

plot(bin1,
     main = "Empirical variogram — log(Al)",
     xlab = "Distance (m)",
     ylab = "Semivariance",
     pch  = 16,
     col  = "darkblue"
)
```

**Interpretation:** The empirical variogram rises from a non-zero intercept at short
distances (a **nugget effect**), increases up to roughly 200,000–300,000 m, and then
appears to level off at a **sill**. This behaviour is consistent with a nested model
comprising a nugget effect and a bounded variogram (e.g. spherical model).

---

# Fitting a Theoretical Variogram Model {#variofit}

We fit a **nested variogram model** consisting of:

- a **nugget effect** $\tau^2$ (pure measurement error / micro-scale variability)
- a **spherical variogram component** with sill $\sigma^2$ and range $\phi$

The full model formula is:

$$
\gamma(h) = \tau^2 + \sigma^2 \cdot \text{Sph}(h;\,\phi)
$$

where $\text{Sph}(h;\,\phi) = \frac{3h}{2\phi} - \frac{1}{2}\left(\frac{h}{\phi}\right)^3$
for $0 < h \le \phi$, and $1$ for $h > \phi$.

```{r variofit, fig.cap="Empirical variogram (dots) with fitted nested model (solid line): nugget + spherical component. Parameters are estimated by weighted least squares."}
# Fit a nested nugget + spherical model to the empirical variogram.
# ini.cov.pars: initial values for c(sill sigma^2, range phi); nugget = tau^2 start
# weights = "npairs": weight each lag class by the number of pairs (more robust)
vfit <- variofit(
  bin1,
  cov.model  = "spherical",
  ini.cov.pars = c(0.15, 250000),  # initial sill and range
  nugget     = 0.05,               # initial nugget
  weights    = "npairs"
)

# Print the fitted parameters
print(vfit)

# Extract and report the key variogram parameters
nugget <- vfit$nugget
sill   <- vfit$cov.pars[1]   # partial sill (sigma^2)
range  <- vfit$cov.pars[2]   # range (phi)

cat("\n--- Fitted variogram parameters ---\n")
cat(sprintf("  Nugget  (tau^2):         %.4f\n", nugget))
cat(sprintf("  Partial sill (sigma^2):  %.4f\n", sill))
cat(sprintf("  Total sill:              %.4f\n", nugget + sill))
cat(sprintf("  Range   (phi):           %.0f m\n", range))
cat(sprintf("  Nugget/Sill ratio:       %.1f %%\n", 100 * nugget / (nugget + sill)))
```

```{r plot-variofit, fig.cap="Fitted nested variogram model (nugget + spherical) overlaid on the empirical variogram."}
# Plot the empirical variogram points and overlay the fitted model line
plot(bin1,
     main = "Variogram model fit — log(Al)",
     xlab = "Distance (m)",
     ylab = "Semivariance",
     pch  = 16,
     col  = "darkblue"
)
# lines.variomodel() draws the theoretical curve from the fitted model object
lines(vfit, col = "red", lwd = 2)
legend("bottomright",
       legend = c(
         "Empirical variogram",
         sprintf("Nugget: %.4f", nugget),
         sprintf("Sill: %.4f", nugget + sill),
         sprintf("Range: %.0f m", range)
       ),
       col = c("darkblue", "white", "white", "white"),
       pch = c(16, NA, NA, NA),
       lty = c(0, 0, 0, 0),
       bty = "n", cex = 0.85
)
lines(vfit, col = "red", lwd = 2)   # re-draw on top of legend box
```

**Interpretation:** The nested model $\gamma(h) = \tau^2 + \sigma^2 \cdot \text{Sph}(h;\phi)$
provides a good fit to the empirical variogram. The nugget effect accounts for
micro-scale variability and measurement error at distances smaller than the sampling
interval. The spherical component captures the spatial autocorrelation structure, reaching
its sill at the practical range $\phi$, beyond which samples can be considered spatially
independent.

---

# Spatial Prediction: Kriging {#kriging}

## Prediction Grid

We set up a regular prediction grid covering the full spatial extent of the data.

```{r pred-grid}
# Determine the spatial extent of the sampling domain
x_range <- range(s100$coords[, 1])
y_range <- range(s100$coords[, 2])
cat("X range (Easting):  ", x_range, "\n")
cat("Y range (Northing): ", y_range, "\n")

# Create a 50×50 regular grid (2,500 prediction locations) across the study area.
# A coarser grid (l=50) is used to keep computation time reasonable;
# increase l (e.g. to 100) for a finer, smoother map.
pred.grid <- expand.grid(
  seq(x_range[1], x_range[2], l = 50),   # 50 equally spaced X values
  seq(y_range[1], y_range[2], l = 50)    # 50 equally spaced Y values
)
cat("Number of prediction grid points:", nrow(pred.grid), "\n")

# Visual check: plot the prediction grid overlaid on sample locations
plot(pred.grid, pch = ".", col = "grey70",
     xlab = "Easting (m)", ylab = "Northing (m)",
     main = "Prediction grid (grey) and sample locations (blue)")
points(s100$coords, pch = 16, col = "steelblue", cex = 0.6)
```

## Kriging Estimation

We use **ordinary kriging** (`krige.conv`) to estimate log(Al) at each grid point.

```{r kriging, cache=TRUE}
# Perform ordinary kriging using the fitted variogram model (vfit).
# krige.conv() applies global neighbourhood (all data used for each prediction).
# For large grids this can be slow — consider local neighbourhood for speed.
kc <- krige.conv(
  s100,
  loc   = pred.grid,
  krige = krige.control(obj.m = vfit)  # use the fitted variogram model
)

cat("Kriging complete.\n")
cat("Predicted values range (log scale):", round(range(kc$predict), 3), "\n")
cat("Back-transformed range (mg/kg):    ", round(range(exp(kc$predict)), 1), "\n")
```

## Kriging Map

```{r kriging-map, fig.cap="Ordinary kriging map of log(Al) across the KOLA study area (50×50 grid). Colour scale: blue = low, red = high log(Al). Sample locations are overlaid as black dots.", fig.height=6}
# image() renders the kriging predictions as a colour-coded raster map.
# The predictions are on the log(Al) scale.
image(kc,
      loc  = pred.grid,
      xlab = "Easting (m)",
      ylab = "Northing (m)",
      main = "Kriging map — log(Al) [KOLA data]"
)

# Overlay the original sample locations so we can relate predictions to data
points(s100$coords, pch = 16, cex = 0.4, col = "black")

# Add a text annotation reminding readers of the scale
mtext("Note: predicted values are on log-scale; exp() for back-transformation",
      side = 1, line = 4, cex = 0.75, col = "grey40")
```

```{r kriging-variance-map, fig.cap="Kriging variance map. Low variance (blue) indicates reliable predictions; high variance (red) indicates uncertainty, typically at the edges or far from sample locations.", fig.height=6}
# The kriging variance kc$krige.var quantifies prediction uncertainty.
# High variance occurs in data-sparse regions (edge effects, gaps in sampling).
image(kc,
      values = kc$krige.var,  # plot variance instead of predictions
      loc    = pred.grid,
      xlab   = "Easting (m)",
      ylab   = "Northing (m)",
      main   = "Kriging variance map — log(Al)"
)
points(s100$coords, pch = 16, cex = 0.4, col = "black")
```

```{r contour-map, fig.cap="Filled contour map of kriging predictions for log(Al). Contour lines help identify spatial patterns and boundaries.", fig.height=6}
# A filled contour plot provides an alternative, often cleaner visualisation.
contour(kc,
        filled      = TRUE,
        coords.data = s100$coords,
        color       = heat.colors,
        values      = kc$predict,
        main        = "Kriging map (contour) — log(Al)",
        xlab        = "Easting (m)",
        ylab        = "Northing (m)"
)
```

**Interpretation:** The kriging map confirms the spatial gradient suggested by the
exploratory analysis: higher Al concentrations in the south-eastern part of the study
area and lower values in the north-west. The kriging variance map shows higher uncertainty
at the boundaries of the study domain where sample density is lower, as expected.

---

# Voluntary Exercise: Sensitivity of the Kriging Map to the Variogram Model {#voluntary}

We examine how changes in the variogram model affect the kriging map by testing four
alternative models:

**(a)** Spherical model *without* a nugget
**(b)** Model (a), but with **half the sill**
**(c)** Model (a), but with **half the range**
**(d)** Pure nugget effect model (no spatial structure)

```{r voluntary-models, cache=TRUE}
# --- Helper function: run kriging with a given model and return predictions ---
krige_with_model <- function(vmod) {
  krige.conv(s100, loc = pred.grid, krige = krige.control(obj.m = vmod))
}

# Base parameters from the fitted model
s2   <- sill           # partial sill
phi  <- range          # range
tau2 <- nugget         # nugget

# (a) Spherical, no nugget: set nugget = 0, keep sill and range
m_a <- variofit(bin1, cov.model = "spherical",
                ini.cov.pars = c(s2 + tau2, phi),  # sill = total sill now
                fix.nugget = TRUE, nugget = 0, weights = "npairs")

# (b) Spherical, no nugget, half the sill
m_b <- variofit(bin1, cov.model = "spherical",
                ini.cov.pars = c((s2 + tau2) / 2, phi),
                fix.nugget = TRUE, nugget = 0, weights = "npairs")

# (c) Spherical, no nugget, half the range
m_c <- variofit(bin1, cov.model = "spherical",
                ini.cov.pars = c(s2 + tau2, phi / 2),
                fix.nugget = TRUE, nugget = 0, weights = "npairs")

# (d) Pure nugget effect: no spatial correlation (sill = 0, large range is irrelevant)
# Represent as exponential with a very small range → converges to pure nugget
m_d <- variofit(bin1, cov.model = "spherical",
                ini.cov.pars = c(1e-6, 1),    # nearly zero spatial variance
                fix.nugget = TRUE, nugget = nugget + sill, weights = "npairs")

# Run kriging for all four alternative models
kc_a <- krige_with_model(m_a)
kc_b <- krige_with_model(m_b)
kc_c <- krige_with_model(m_c)
kc_d <- krige_with_model(m_d)

cat("Alternative models computed successfully.\n")
```

```{r voluntary-plot, fig.cap="Kriging maps for four alternative variogram models. (a) No nugget — smoothest map. (b) No nugget, half sill — same pattern, narrower colour range. (c) No nugget, half range — sharper local features. (d) Pure nugget — spatially uncorrelated, map collapses to global mean.", fig.height=6, fig.width=6.5, out.width="\\textwidth"}
# Plot all four alternative maps side by side for comparison
par(mfrow = c(2, 2), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.6, 0))

image(kc_a, loc = pred.grid, xlab = "Easting", ylab = "Northing",
      main = "(a) Spherical, no nugget")
points(s100$coords, pch = ".", cex = 0.5)

image(kc_b, loc = pred.grid, xlab = "Easting", ylab = "Northing",
      main = "(b) No nugget, half sill")
points(s100$coords, pch = ".", cex = 0.5)

image(kc_c, loc = pred.grid, xlab = "Easting", ylab = "Northing",
      main = "(c) No nugget, half range")
points(s100$coords, pch = ".", cex = 0.5)

image(kc_d, loc = pred.grid, xlab = "Easting", ylab = "Northing",
      main = "(d) Pure nugget effect")
points(s100$coords, pch = ".", cex = 0.5)

par(mfrow = c(1, 1))  # reset layout
```

**Interpretation of the sensitivity analysis:**

- **(a) No nugget:** Removing the nugget makes the kriging interpolation perfectly honour
  the data values at sample locations (exact interpolation), producing a slightly
  rougher map with more pronounced local extremes.

- **(b) Half sill:** Halving the sill reduces the total variability assumed in the model.
  The spatial pattern remains identical to (a) but the colour range narrows — the map
  appears less contrasted, reflecting that the model now assumes less spatial variance.

- **(c) Half range:** Reducing the range makes the model assume that spatial correlation
  dies out at shorter distances. Kriging relies less on distant samples, producing
  sharper, more localised features and more rapid transitions between high and low values.

- **(d) Pure nugget:** A pure nugget model assumes *no* spatial correlation — all
  variability is random. Kriging with this model effectively returns the global mean
  everywhere (with no smoothing), resulting in a nearly uniform map. This demonstrates
  why the variogram model matters: ignoring spatial structure yields uninformative maps.

---

# Summary

| Analysis step         | Result / Decision |
|-----------------------|------------------------------------------------|
| Transformation        | Log-transformation applied (right-skewed data) |
| Nugget $\tau^2$       | `r round(nugget, 4)` |
| Partial sill $\sigma^2$ | `r round(sill, 4)` |
| Total sill            | `r round(nugget + sill, 4)` |
| Range $\phi$          | `r round(range, 0)` m |
| Nugget/Sill ratio     | `r round(100 * nugget / (nugget + sill), 1)` % |

The geostatistical analysis reveals a clear spatial structure in Al concentrations across
the KOLA study area. The fitted nested variogram model (nugget + spherical) provides a
good description of the spatial autocorrelation. The kriging map highlights a large-scale
gradient, likely driven by geological background variation in the Fennoscandian bedrock.
