---
title: "Environmental Statistics"
author: "Strachinaru Andrei Lucian"
matrikel_number: "#####"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: united
    highlight: haddock
    toc: yes
    toc_float: yes
    fig_width: 8  
    fig_height: 5
    df_print: paged
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  paged.print = FALSE,
  echo        = TRUE,
  fig.show    = "asis",      # ← ensures plots are shown inline
  fig.width   = 8,
  fig.height  = 5,
  warning     = FALSE,
  message     = FALSE
)
knitr::opts_knit$set(root.dir = "c:/Users/andrei.strachinaru/OneDrive/BOKU/S2026/Environmental_Statistics/5_Exercises/Exercises/Exercise_2_files/")
```

# Exercise 2

## Exercise 2.1

Plot the hydrograph for gauge Ibm/Moosache for the year 1976. Mark all extreme values above the 0.95 quantile with a red asterisk.

```{r ex_2.1}

#setting up the file
q1 <- as.numeric(readLines("file0.txt"))

cat("Lenght of data is: ", length(q1), ".", sep="")

#define year 1976
year_1976 <- q1[1:366]
cat("Number of days of year 1976 is:", length(year_1976), "days.")

#define the quantile q95_1976 and store it
q95_1976 <- quantile(year_1976, probs = 0.95)
cat("The threshold for quantile q95_1976 is: ", q95_1976," m³/s.", sep="")

#define the values that overpasses the required quantile
extreme_values_1976 <- which(year_1976 > q95_1976)
cat("The values above the threshold for year 1976 are:", year_1976[extreme_values_1976])

#plot the Discharge of year 1976
print(plot(year_1976,
    type = "l"))

#plot the points that are above q95_1976
points(extreme_values_1976, year_1976[extreme_values_1976],
    pch = "*")

```

## Exercise 2.2

Explore the graphical options of R to make the graph appealing. Produce a complete graph
“ready for publishing” which gives the observer all necessary information, i.e. do not forget
correct axes labels (with units!), title, etc.

```{r ex_2.2}
#plot the Discharge of year 1976
plot(year_1976,
    type = "l",
    lwd = 2, #bolding the line
    lty = 1, #dashing the line
    col = "lightblue", #coloring the line
    xlab = "Days of year 1976",
    ylab = "Discharge (m³/s)",
    main = "Pegel Ibm/Moosache - Discharge gauge",)


#define the maximum line at quantile q95_1976
abline(h = q95_1976, 
    col = "red",
    lwd = 2,
    lty = 2)
text(1, q95_1976, 
    "Quantile 0.95",
    col = "red",
    adj = c(-1, -1) )

#plot the points
points(extreme_values_1976, year_1976[extreme_values_1976],
    pch = "*",
    col = "red",
    lwd = 2)

legend("topright",
    legend = c("Discharge", 
                  paste0("0.95 Quantile (", round(q95_1976, 3), " m³/s)"),
                  "Extreme values"),
    col    = c("steelblue", "red", "red"),
    lty    = c(1, 2, NA),
    pch    = c(NA, NA, 8),
    lwd    = c(1.5, 2, 2),
    y.intersp = 1.2)
```

## Exercise 2.3

Produce a similar plot for 1977.

```{r ex_2.3}
#defining year 1967 form the file
year_1977 <- q1[367:732]
cat("Number of days of year 1977 is:",length(year_1977), "days.")

#define the quantile q95_1976 for year 1977 and store it in a variable
q95_1977 <- quantile(year_1977, probs = 0.95)
cat("The threshold for quantile q95 for year 1977 is: ", q95_1977, " m³/s.", sep="")

#define the extreme values that overpass q95_1977
extreme_values_1977 <- which(year_1977 > q95_1977)
cat("The values above the threshold for year 1976 are:", year_1977[extreme_values_1977], ".")

#plot the Discharge of year 1976
plot(year_1977,
    type = "l",
    lwd = 2, #bolding the line
    lty = 1, #dashing the line
    col = "lightblue", #coloring the line
    xlab = "Days of year 1977",
    ylab = "Discharge (m³/s)",
    main = "Pegel Ibm/Moosache - Discharge gauge",)


#define the maximum line at quantile q95_1976
abline(h = q95_1977, 
    col = "red",
    lwd = 2,
    lty = 2)
text(1, q95_1977, 
    "Quantile 0.95",
    col = "red",
    adj = c(-5, -1.5) )

#plot the points
points(extreme_values_1977, year_1977[extreme_values_1977],
    pch = "*",
    col = "red",
    lwd = 2)

legend("topright",
    legend = c("Discharge", 
                  paste0("0.95 Quantile (", round(q95_1977, 3), " m³/s)"),
                  "Extreme values"),
    col    = c("steelblue", "red", "red"),
    lty    = c(1, 2, NA),
    pch    = c(NA, NA, 8),
    lwd    = c(1.5, 2, 2),
    y.intersp = 1.2)

```