---
title: "Environmental Statistics"
author: "Strachinaru Andrei Lucian"
matrikel_number: "#####"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: united
    highlight: haddock
    toc: yes
    toc_float: yes
    fig_width: 8  
    fig_height: 5
    df_print: paged
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(paged.print = FALSE)
```

# Exercise 2

## Exercise 1.1

Calculate the first 50 power values of 2, i.e., **$2^{1}=1, 2^{2}=4, ...$**
. Then calculate the squares of all integer numbers from 1 to 50, i.e. **$1^{2}=1∗1=1, 2^{2}=2∗2=4, ...$**
 and so on. Which of the number pairs fulfill the condition **$2^{n}=n*{2}$**
?

```{r ex_1.1}
#define p, s as a 50 dimensional vector
p <- numeric(50)
s <- numeric(50)
#iterate from 1 to 50 and calculate p and s
for (x in 1:50){
  p[x] <- 2 ^ x
  s[x] <- x ^ x
}
#print p
print(p)
#print s
print(s)
#iterate for all the values and verify the if and print the conditions
for (x in 1:50){
  if( 2 ^ x == x * 2){
    cat("x =", x, "| 2 ^", x, "==", x, "* 2", "\n")
  }
}

```

## Exercise 1.2

For the numbers 0, 0.1, 0.2, …, 2π calculate sine, cosine and tangent using the R functions sin(), cos(), tan(). Alternatively, calculate tangent using the equation tan(x)=sin(x)/cos(x). Compare both calculation methods for the tangent. How many numbers are equal, and what is the largest difference between both calculations?

```{r ex_1.2}
#generate numbers that hav 0.1 increments from 0 to 2 * pi
x <- seq(0, 2 * pi, by = 0.1)
#definie a vector that contains the absolute values of the difference between tan and sin/cos
tan_diff <- abs(tan(x) - sin(x)/cos(x))
#define a data.frame with x, sine, cousine, and the both tangents that are calculated using different methods
results <- data.frame(
  x = round(x, 3),
  sine = round(sin(x), 3), #radians
  cosine = round(cos(x), 3), #radians
  tang_r = tan(x), #radians
  tang_div = sin(x)/cos(x), #radians
  tan_diff #radians
)
#display the results
print(results)
#display the maximum value of the difference between both methods used to calculate the tangent
cat("Max difference is: ", max(tan_diff), " radians.")
#second method to calculate and display the maximum difference between both tangent methods
difference <- abs(results$tang_r - results$tang_div)
max_difference <- max(difference)
cat("Max difference between tang_r and tang_div is", max_difference, " radians.")

#compare the results and store the results that are equal into a vector compare
compare <- results$tang_r == results$tang_div
#add the values to count the number of times the condition of equality is fulfilled
count_equal <- sum(compare)
cat("There are", count_equal, "out of", length(x), "values that are equal.")


```

## Exercise 1.3

Here we use the dataset ngaruroro of the package lfstat, which contains daily streamflow data from an Australian stream gauge. Use install.packages(), library(), and data() to install the package from CRAN and load the data. Then inspect the dataset using the str(), head(), tail() and class() commands. Finally perform the following tasks:

- *Select the flow record of 1999 (and store it as an own object).*
- *Plot the flow record of 1999 (hint: maybe the generic function plot() will be so clever to offer a dedicated plot for the lfstat data class).*
- *What is the mean flow of the year 1999?*
- *On which day was the annual maximum flow recorded in 1999?*

```{r ex_1.3, warning=FALSE, message=FALSE}

#install packages if they are missing
if (!require("lfstat")) install.packages("lfstat")
if (!require("DT")) install.packages("DT")

#access libraries
library(lfstat)
library(DT)

#access the dataset named "ngaruroro"
data(ngaruroro)
#assign the dataset to "streamflow"
streamflow <- ngaruroro

#display the first 10 rows as a table using knitr
knitr::kable(head(streamflow, 10))
#another way to display all the data inro an R Markdown file without agglomarate it
DT::datatable(streamflow)

#assign to flow_1999 all the data from the year 1999
flow_1999 <- streamflow[streamflow$year == 1999, ]

#plot the data without assigning anything to x and y
plot(flow_1999)

#calculate mean flow
mean_flow <- mean(flow_1999$flow, na.rm = "TRUE")
cat("Mean flow value is ", round(mean_flow, 4), " m³/s .", sep="")

#calculate max flow
max_flow <- max(flow_1999$flow)
#identify the day when the max flow was recorded
max_day_index <- which.max(flow_1999$flow)
max_day <- flow_1999$day[max_day_index]
#display the message
cat("The annual maximum flow recorded on ", flow_1999$day[max_day_index],".",flow_1999$month[max_day_index],".", 
flow_1999$year[max_day_index]," was ", flow_1999$flow[max_day_index], " m³/s.",sep = "")

```
